Modelling last glacial cycle ice dynamics in the Alps

### 2017-09-25 articles/cycle/alps-cycle.tex
Rephrase glacier dynamics section.

    During MIS~2, all six simulations yield comparative cumulative glaciated
    areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
    palaeo-climate forcing approach used, which involved a linear scaling of
    each palaeo-temperature anomaly record to a geomorphological
    reconstruction of the area glaciated by the Rhine Glacier piedmont lobe
    (Sect.~\ref{sec:paltemp}, Fig.~\ref{fig:footprints}a, black rectangle).
    Concerning other areas, all simulations underestimate glaciated area in the
    western part of the model domain, and overestimate it in the western
    part, relative to the geomorphological ice limits
    (Fig.~\ref{fig:footprints}a). This result might indicate that the LGM
    temperature depression, here considered homogeneous, was actually lower in
    the Eastern Alps and in the Western Alps. Alternatively, the east-west
    gradient in precipitation existing today (Fig.~\ref{fig:inputs}d) was
    perhaps enhanced during the LGM, as indicated by pollen reconstruction
    \citep{Wu.etal.2007} and positive degree-day modelling of central European
    palaeo-glaciers \citep{Heyman.etal.2013}.

@@

    The LGM extent of alpine glaciers has been mapped with varying level of
    detail across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
    Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
    Buoncristiani.Campy.2011, Husen.2011}. These maps have been compiled
    into an reconstruction covering the entire Alps \citep{Ehlers.etal.2011},
    reproduced here (Fig.~4, red line) for comparison against model results.

    The modelled total ice volume reaches a maximum of $122.5 \times 10^{3}$\,\unit{km^2},
    or 300.4\,mm of sea-level equivalent, at 24.56\,ka
    (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of
    $163.3 \times 10^{3}$\,\unit{km^2} is attained shortly aftewards at 24.57\,ka.
    (Fig.~\ref{fig:lgmvel}b). Although the modelled timing of the LGM varies
    accross the mountain range (Sect.~\ref{sec:timing}), at 24.57\,ka nearly
    all glaciers extend to within a few km from their modelled maximum stage
    (Fig.~\ref{fig:lgmvel}a, dashed orange line).

    Although the palaeo-climate forcing was adapted (Sect.~\ref{sec:climate})
    to model this maximum configuration in broad agreement with the geological
    reconstruction of the LGM extent (Fig.~\ref{fig:lgmvel}a, solid red line),
    local discrepancies remain. Most notably, reconstructed ice extent in the
    north-western Alps, including the Rhone Glacier complex, the Jura ice cap,
    and the Lyon Lobe, is largely missing from our model results. On the other
    hand, the model yields excessive glaciation in the eastern Alps, where the
    Palten, Drau and Sava glaciers are modelled to extend tens of kilometers
    beyond the mapped moraines. These discrepancies might indicate that the
    LGM climate was characterized by an east-west gradient in temperature or
    precipitation anomalies relative to present not included in the climate
    forcing, or perhaps a east-west gradient in variables not accounted for
    by the PDD model, such as cloud cover or dust deposition.

    On a more regional level, the modelled maximum extent of the Ticino, Iseo,
    Garda and Piave glaciers in the central southern Alps is constrained
    several kilometres within the mapped LGM moraines. This appears to result
    from the paleo-precipitation reduction used to force the model
    (Fig.~\ref{fig:footprints}b), which is likely unrealistic in at least this
    part of the model domain. On the other hand, the Durance glacier in the
    south-western Alps is modelled to extend several kilometers beyond the
    mapped limit, indicating that the LGM temperature depression was likely
    dampened by Mediterranean climate in this part of the model domain.

    The model reproduces peripheral ice caps where documented by geological
    evidence on the Vercors, Chartreuse and Bauge prealpine reliefs,
    the Jura Mountains, the Vosges Mountain, the Black Forest, the Bohemian
    Forest and the Dinaric Alps. An independent ice cap also covers the
    Hochwart massif during most of the simulation, yet it is engulfed by
    alpine glaciers during the LGM to become a peripheral ice dome. The LGM
    extent of peripheral ice caps is underestimated in the Vosges and Jura
    mountains, while it is overestimated for the Vercors. There is a conformity
    between the discrepancies between modelled and reconstructed LGM extent
    for the main alpine glacier complex vs peripheral ice caps, which indicates
    a climatic cause rather than an ice dynamics cause.

    \todo{add location names on Fig.~\ref{fig:lgmvel}.}

@@

    The LGM Alpine ice flow pattern was traditionally described as that of a
    network of interconnected valley glaciers that was primarily controlled by
    subglacial topography. However, the geomorphology shows that ice was thick
    enough to flow accross high mountain passes throughout the mountain range
    \citep[e.g.,][]{Onde.1938, Penck.Bruckner.1909, Jackli.1962, Husen.1985,
    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even perhaps form
    self-sustained ice domes \citep{Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009}.

    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a).
    The modelled fast-flow regions generally occur along the main river
    valleys, while ice domes and ice divides are predominantly located over
    major reliefs . Nevertheless, the model results depict occasional ice flow
    across mountain passes (transfluences) and ice divide above topographic
    lows.

    Major transfluences occur for instance across col du Mont-Cenis,
    Simplonpass, Brünigpass in the western Alps, and over Fernpass,
    Gailbergsattel and Stranigeralpe in the Eastern Alps
    (Fig.~\ref{fig:lgmvel}a). Finally, self-sustained ice domes characteristic
    of ice caps are modelled in two locations over Flüelapass and Ötztal
    (Fig.~\ref{fig:lgmvel}a).

    Transfluence across col de Montgenèvre was previously questioned
    \citep[Fig.~2]{Cossart.etal.2012}. Transfluence across col de Mont Cenis
    was previoysly described \citep[Fig.~3.18]{Onde.1938, Coutterand.2010}.
    Transfluences across Fernpass and Seefeld are known from the geomorphology
    \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. Transfluence across
    Brunigpass was already marked in the maps \citep{Jackli.1962}. There is
    numerous evidence for transfluence across Simplon Pass
    \citep{Kelly.etal.2004}. Ice crossed the Pyhrn Pass \citep{Husen.2004}. Ice
    flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}. The
    Tagliamento catchment is usually assumed to not have received tranfluences
    from the Drau catchment \citep{Monegato.etal.2007}, but this would not be
    incompatible with reconstructed ice surface elevations in this area
    \citep{Husen.1987}.

    In the Eastern Alps, modelled transfluences and crosswise divides are
    generally incompatible with geological reconstructions, indicating that
    climate deterioration was overestimated in the model input. In the Western
    Alps, however, transfluences generally occur where they have been
    documented by geologic evidence. These model results depict the LGM Alpine
    ice complex as an intermediate between ice fields and ice caps, bearing
    characteristics of both and perhaps no modern analogue.

    \todo{add location names on Fig.~\ref{fig:lgmvel}.}

%    \citep{Jouvet.etal.2017}

@@

    The timing of the LGM has been documented by radiocarbon and cosmogenic
    isotope dating techniques at multiple locations around the Alps (cf.
    \citealp[Fig.~5]{Wirsig.etal.2016} for a review and the more recent
    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
    data indicate that Alpine glaciers reached their maximum extent between 20
    and 26\,ka (calibrated \chem{^{14}C} or \chem{^{10}Be} ages), but glacier
    lobes stayed in the foreland until as late as 22 to 17\,ka, when they
    experienced rapid retreat synchronously to lowering of the ice surface in
    the mountains (\citealp[Fig.~5]{Wirsig.etal.2016},
    \citealp[Fig.~3]{Monegato.etal.2017}).

@@

    It is important to notice that these differences in timing of the LGM occur
    in the model results despite the simple climate forcing used consisting of
    homogeneous anomalies. These discrepancies between different glacier lobes
    are thus not related to climate, but are a pure product of modelled
    inherent glacier dynamics. They occur because of differences in the
    subglacial topography, in particular catchment sizes and bed roughness.

    PISM is a polythermal ice sheet model, i.e. it computes the distribution
    of englacial temperatures resulting from heat advection, vertical
    diffusion in ice and rock, and strain heat release, in three dimensions.
    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
    Rhine and Garda glaciers, several thousands of years are needed to attain
    a thermodynamical equilibrium between cold ice advection from the high
    accumulation areas, upward diffusion of geothermal heat, and heat release
    from strong basal shear strain. Temperature evolution is, in part, limited
    by the slow warming of subglacial bedrock, that has been previously cooled
    downed by subfreezing air temperature before glacier advance.

    For such glaciers, this thermodynamical equilibrium is typically not yet
    attained around 25\,ka. This is why several Alpine lobes overshoot their
    balanced extent before thinning and receding towards the mountains as they
    warm towards thermodynamical equilibrium.

    Our results indicate that even for a relatively small ice sheet like that
    found in the Alps, differences in timing on the order of the millenium can
    result from complex glacier dynamics. Further differences in timing of the
    LGM caused by heterogeneous climatic anomalies, such as precipitation
    anomaly gradients, not included in our model set-up, may have
    counterbalanced or emphasize those presented here.

    \todo{add location names on Fig.~\ref{fig:lgmvel}.}

@@

    Following the general assumption that trimlines, which mark the transition
    between the steep frost-shattered ridges and the more gentle glacially eroded
    valley troughs, represent the maximum elevation of the LGM ice surface,
    the LGM ice thickness in the Alps been reconstructed in several areas
    \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
    However, this assumption is challenged by geological evidence from
    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994,
    Stroeven.etal.2002}, the British Isles \citep[e.g.,][]{Fabel.etal.2012}, and
    North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
    located well above the trimlines have been glaciated and preserved under
    under cold-based ice, sometimes for several glacial cycles.

    In several places across the Alps, the maximum elevation reached by the
    ice surface during the LGM has been reconstructed from trimlines, the upper
    limit of glacial erosion observed in the landscape. This include the entire
    western Alps, the Rhone Glacier catchment, the Rhine Glacier catchment,
    south-eastern Switzerland and all of Austria. These reconstructions are
    based on the assumption that glacial ice was significantly erosive up to
    the or almost up to the surface, which was then located at or slightly
    above the elevation of the trimlines. This assumption is supported by the
    fact that there exist a regional consistence in trimlines elevations, which
    are generally located along a surface sloping in the assumed direction of
    flow. However, in other glaciated mountain regions, including several
    sites in the Scandinavian Mountains, the Cairgorns in Scotland, and high
    plateaus in the Canadian Cordillera, that cold-based ice commonly occured
    at high elevation and preserved the pre-glacial landscape.

    In the upper Rhone valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a, b), with a mean
    difference of 861\,m (xx\,\%, Fig.~\ref{fig:trimlines}b). This result is
    somewhat depedent on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordillera ice sheet show that varying ice rheological parameters
    resulted relative errors of 25\,\unit{\%} with regard to ice rheological
    parameters and 14\,\unit{\%} regarding basal sliding parameters
    \citep[Fig.~7]{Seguinot.etal.2016}.

    In addition, our model results depict a polythermal LGM Alpine ice cover.
    Due to sub-freezing temperatures applied in the climate forcing of the
    model, the entire Alpine ice sheet is capped by an upper layer of cold ice.
    In major Alpine valleys, such as the upper Rhone Valley, important ice
    thickness and strain heating contribute to form a layer of temperate ice
    near the glacier base, allowing basal melt, sliding and potentially
    erosion. On the other hand, on the highest mountains ice cover is to thin
    and too static to form temperate ice, resulting in cold ice down to the
    bed, preventing potential erosion.

    In the upper Rhone valley, the geographical location of observed trimlines
    show remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based for less than 1\,ka (7\,\%) of the
    MIS~2. Although this result calls for more detailed comparisons across the
    entire Alpine range and specific sensitivity studies to relevant basal
    sliding and ice rheological parameters, the presence of an upper layer of
    cold ice during the LGM, already found in some places in a much warmer
    climate today, is inevitable. Thus we want to challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer cold ice in the Alps.

    \todo{look for references.}

@@

    As previously mentioned (Sect.~\ref{sec:timeseries}), independently of the
    palaeo-temperature (GRIP, EPICA, or MD01-2444) and palaeo-precipitation
    (with or without corrections) applied, all simulations presented here
    result in a high time variability in the total modelled ice volume
    (Fig.~\ref{fig:timeseries}). Unsurprisingly, this result is confirmed by
    the 1-km resolution run, using the optimal EPICA palaeo-temperature record
    (Sect.~\ref{sec:paltemp}) and conservatively includes palaeo-precipitation
    reductions (Sect.~\ref{sec:palprec}), which also results in strong total
    ice volume variability throughout the last glacial cycle
    (Fig.~\ref{fig:lgmvel}b).

    Two major glaciations occur during MIS~4 and 2 (Fig.~\ref{fig:lgmvel}b).
    However, several minor glaciations occur during MIS~5 and 3, as well as a
    minor late-glacial readvance at the onset of MIS~1
    (Fig.~\ref{fig:lgmvel}b). These episodes are the result of synchronous
    advances several Alpine glaciers well into the major valleys and sometimes
    even onto the foreland (Fig.~\ref{fig:profiles}), Supplementary animation).

@@

    Despite the low temperature variability of the palaeo-climate forcing, and
    reduced precipitation dampening glacier response, our simulation depict
    the Alpine ice complex as highly dynamic, with dynamic of individual
    glaciers controlled by variations in subglacial topography and catchment
    geometry.

    \idea{Matthias: The selection of figures in the main text and SOM is good.
          I find Figure 5 quite difficult to understand at first glance. I
          would combine it (already before submission) with Fig. 6. the Figure
          can then focus both on the temporal advance/retreat patterns of the 4
          lobes (four panels on the left side of figure) and the spatial
          differences in max. ice elevation (1 panel on right side of figure,
          with centerlines indicated). I would have the one with trimline
          elevation again in the Supplementary.}

>>>>

    During MIS~2, all six simulations yield comparative cumulative glaciated
    areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
    palaeo-climate forcing approach used, which involved a linear scaling of
    each palaeo-temperature anomaly record to a geomorphological
    reconstruction of the area glaciated by the Rhine Glacier piedmont lobe
    (Sect.~\ref{sec:paltemp}, Fig.~\ref{fig:footprints}a, black rectangle).
    Concerning other areas, all simulations underestimate glaciated area in the
    western part of the model domain, and overestimate it in the western
    part, relative to the geomorphological ice limits
    (Fig.~\ref{fig:footprints}a). This result might indicate that the LGM
    temperature depression, here considered homogeneous, was actually lower in
    the Eastern Alps and in the Western Alps, as previously shown by positive
    degree-day modelling of the central European palaeo-ice caps
    \citep{Heyman.etal.2013}. Alternatively, the east-west
    gradient in precipitation existing today (Fig.~\ref{fig:inputs}d) was
    perhaps enhanced during the LGM, as indicated by pollen reconstruction
    \citep{Wu.etal.2007}. Finally, the LGM extent of Alpine glaciers might have
    been affected by east-west gradient in variables not accounted for
    by the PDD model, such as cloud cover or dust deposition.

@@

    The LGM extent of alpine glaciers has been mapped with varying level of
    detail across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
    Husen.1987, Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
    Buoncristiani.Campy.2011, Husen.2011}. These maps have been compiled
    into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
    and reproduced here (Fig.~4, red line) for comparison against model
    results.

    The modelled total ice volume reaches a maximum of $122.5 \times
    10^{3}$\,\unit{km^3}, or 300.4\,mm of sea-level equivalent, at 24.56\,ka
    (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of $163.3 \times
    10^{3}$\,\unit{km^2} is attained shortly aftewards at 24.57\,ka
    (Fig.~\ref{fig:lgmvel}b). Although the modelled timing of the LGM varies
    accross the mountain range (Sect.~\ref{sec:timing}), at 24.57\,ka nearly
    all outlet glaciers extend to within a few km from their modelled maximum
    stage (Fig.~\ref{fig:lgmvel}a, dashed orange line).

    The palaeo-climate forcing was adapted (Sect.~\ref{sec:paltemp})
    to model this maximum configuration in broad agreement with the geological
    reconstruction of the LGM extent (Fig.~\ref{fig:lgmvel}a, solid red line),
    yet local discrepancies remain. As already outlined for low resolution
    runs (Sect.~\ref{sec:footprints}), the extent of glaciation in the
    north-western Alps, including the Rhone Glacier complex, the Jura ice cap,
    and the Lyon Lobe, is underestimated in the model results
    (Fig.~\ref{fig:lgmvel}a). On the other hand, the model yields excessive ice
    cover in the eastern Alps, where the Palten, Drau and Sava glaciers are
    modelled to extend tens of km beyond the mapped end-moraines
    (Fig.~\ref{fig:lgmvel}a). As previously discussed
    (Sect.~\ref{sec:footprints}), these discrepancies might indicate that the
    LGM climate was characterized by an east-west gradient in temperature
    \citep[cf.][]{Heyman.etal.2013} or precipitation
    \citep[cf.][]{Wu.etal.2007} anomalies relative to present, or an east-west
    gradient in variables not accounted for by the PDD model.

    On a more regional level, the modelled maximum extent of the Ticino, Iseo,
    Garda and Piave glaciers in the central southern Alps is constrained
    several kilometres within the mapped LGM moraines (Fig.~\ref{fig:lgmvel}a).
    This appears to result
    from the paleo-precipitation reduction used to force the model
    (Fig.~\ref{fig:footprints}b), which is likely unrealistic in at least this
    part of the model domain. On the other hand, the Durance glacier in the
    south-western Alps is modelled to extend several kilometers beyond the
    mapped limit (Fig.~\ref{fig:lgmvel}a), indicating that the LGM temperature
    depression was likely
    dampened by Mediterranean climate in this part of the model domain.

    The model reproduces peripheral ice caps where documented by geological
    evidence on the Vercors, Chartreuse and Bauge prealpine reliefs,
    the Jura Mountains, the Vosges Mountain, the Black Forest, the Bohemian
    Forest and the Dinaric Alps (Fig.~\ref{fig:lgmvel}a). An independent ice
    cap also covers the Hochwart massif during most of the simulation, yet it
    is engulfed by alpine glaciers during the LGM to become a peripheral ice
    dome \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}. The LGM
    extent of peripheral ice caps is underestimated in the Vosges and Jura
    mountains, while it is overestimated for the Vercors
    \citep[Fig.~\ref{fig:lgmvel}a; cf.][Figs.~4.28, 4.32, and
    4.33]{Coutterand.2010}. Thus, there exists a regional conformity between
    the model-data discrepancies obtained for the main Alpine ice sheet and
    those obtained for peripheral ice caps. Because peripheral ice cap have
    very different glacier dynamics than the main ice sheet, this conformity
    most likely indicates a climatic cause, rather than an ice dynamics cause,
    for these discrepancies.

    \todo{add location names on Fig.~\ref{fig:lgmvel}: Rhone Glacier, Jura ice
          cap, and the Lyon Lobe, Palten, Drau and Sava glaciers, Ticino, Iseo,
          Garda and Piave glaciers, Durance glacier, Vercors, Chartreuse,
          Bauge, Jura, Vosges, Black Forest, Bohemian Forest, Dinaric Alps,
          Hochwart.}

@@

    The LGM Alpine ice flow pattern was traditionally described as that of a
    network of interconnected valley glaciers, primarily controlled by
    subglacial topography. However, the geomorphology shows that ice was thick
    enough to flow accross high mountain passes throughout the mountain range
    \citep[e.g.,][]{Onde.1938, Penck.Bruckner.1909, Jackli.1962, Husen.1985,
    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even to perhaps form
    self-sustained ice domes \citep{Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009}.

    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a,
    blue colour mapping). Fast-flow regions generally occur along the main
    river valleys, while ice domes and ice divides are predominantly located
    over major reliefs (Fig.~\ref{fig:lgmvel}a). Nevertheless, the model
    results depict occasional ice flow across the modern water divides, i.e.
    transfluences.

    In the Western Alps, major transfluences occur for instance across Col de
    Montgenèvre, Col du Mont-Cenis, Simplon Pass and Brünig Pass
    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across col de Montgenèvre
    was previously questioned \citep[Fig.~2]{Cossart.etal.2012}, evidence for
    southerly ice flow across Mont-Cenis has been recognized
    \citep[Fig.~3.18]{Onde.1938, Coutterand.2010}. Similarly, tranfluences have
    been previously identified from the geomorphology across Simplon Pass
    \citep{Kelly.etal.2004} and Brünig Pass \citep{Jackli.1962}.

    In the Eastern Alps, major transfluences are modelled to have occur for
    instance across at Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
    Kreuxberg Saddle, the Straniger Alp, and the Pyhrn Pass
    (Fig.~\ref{fig:lgmvel}a). Transfluences across Fern Pass and the Seefeld
    Saddle are known from the geomorphology
    \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. It is also known that
    ice flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}, and
    the Pyhrn Pass \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}.
    On the other hand, no transfluence has previously been documented at the
    Straniger Alp. The Tagliamento catchment is usually assumed to not have
    received tranfluences from the Drau catchment \citep{Monegato.etal.2007},
    but this would not be incompatible with reconstructed ice surface
    elevations in this area \citep{Husen.1987}.

    Finally, self-sustained ice domes characteristic of ice caps are modelled
    in two locations over Flüelapass and Ötztal (Fig.~\ref{fig:lgmvel}a).
    Except for too extensive ice cover in the easternmost part of the model
    domain (Sect.~\ref{sec:extent}), there is generally a good agreement
    between transfluences observed in the geomorphology and the model results.
    Both depict the LGM Alpine ice flow pattern as an intermediate between that
    of a topography-controlled ice field, and that of a self-sustained ice cap.

    \todo{add location names on Fig.~\ref{fig:lgmvel}: Mont-Cenis, Simplonpass,
          Brünigpass, Fernpass, Seefeld, Gailbergsattel and Stranigeralpe,
          Tagliamento, Drau, Flüelapass, Ötztal.}

@@

    The timing of the LGM has been documented by radiocarbon and cosmogenic
    isotope dating techniques at multiple locations around the Alps (cf.
    \citealp[Fig.~5]{Wirsig.etal.2016} for a review and the more recent
    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
    data indicate that Alpine glaciers reached their maximum extent between 20
    and 26\,ka (calibrated \chem{^{14}C} and \chem{^{10}Be} ages), but also
    that terminal lobes stayed in the foreland until as late as 22 to 17\,ka,
    when they experienced rapid retreat synchronously to lowering of the ice
    surface in the mountains (\citealp[Fig.~5]{Wirsig.etal.2016},
    \citealp[Fig.~3]{Monegato.etal.2017}).

@@

    These differences in timing of the LGM occur in the model results despite
    the homogeneous temperature and precipitation anomalies supplied as
    palaeo-climate forcing used. The LGM timing differences modelled here are
    thus not related to climate, but are inherent to modelled glacier dynamics.
    They result from differences in the subglacial topography, in particular
    catchment sizes and bed roughness, of the different outlet glaciers.

    PISM is a polythermal ice sheet model, i.e. it computes the distribution
    of englacial temperatures resulting from heat advection, vertical
    diffusion in ice and rock, and strain heat release, in three dimensions.
    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
    Rhine and Garda glaciers, several thousand model years are needed to attain
    a thermodynamical equilibrium between cold ice advection from the high
    accumulation areas, upward diffusion of geothermal heat, and heat release
    from strong basal shear strain. Temperature evolution is, in part, limited
    by the slow warming of subglacial bedrock, that has been previously cooled
    downed by subfreezing air temperature before glacier advance.
    For larger glaciers, this thermodynamical equilibrium is typically not yet
    attained around 25\,ka. This is why several Alpine lobes overshoot their
    balanced extent before thinning and receding towards the mountains as they
    warm towards thermodynamical equilibrium.

    Our results indicate that even for a relatively small ice sheet like that
    found in the Alps, ka-scale differences in timing of the LGM can
    result from complex glacier dynamics. Further differences in LGM timing
    caused by heterogeneous climatic anomalies, such as precipitation
    anomaly gradients, not included in our model set-up, may have
    counterbalanced or emphasize those modelled here.

    \todo{add location names on Figs.~\ref{fig:lgmvel} and \ref{fig:timing}:
    Dora Riparia, Ivrea, and Tagliamento, Durance, Rhone, Inn, Enns, Ticino,
    and Garda, Isère, Como and Iseo, Vercors, Jura Mountains, Black Forest and
    Hochschwab.}

@@

    Following the general assumption that trimlines, which mark the transition
    between the steep frost-shattered ridges and the more gentle glacially eroded
    valley troughs, represent the maximum elevation of the LGM ice surface,
    maximum ice thickness in the Alps been reconstructed in several areas
    \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
    However, this assumption is challenged by geological evidence from
    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994},
    the British Isles \citep[e.g.,][]{Fabel.etal.2012}, and
    North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
    located well above the trimlines have been glaciated and preserved under
    under cold-based ice, sometimes for several glacial cycles
    \citep{Stroeven.etal.2002}.

    In the upper Rhone valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (xx\,\%, Fig.~\ref{fig:trimlines}b). This result is
    somewhat depedent on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordillera ice sheet show that varying ice rheological parameters
    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}.

    Beside, our model results depict a polythermal LGM Alpine ice cover.
    Due to sub-freezing temperatures applied in the climate forcing of the
    model, the entire Alpine ice sheet is capped by an upper layer of cold ice.
    In major Alpine valleys, such as the upper Rhone Valley, important ice
    thickness and strain heating contribute to form a layer of temperate ice
    near the glacier base, allowing basal melt, sliding and potentially
    erosion (Fig.~\ref{fig:trimlines}c, white areas). On the other hand, on the
    highest mountains ice cover is to thin and too static to form temperate
    ice, resulting in cold ice down to the bed, preventing potential erosion
    (Fig.~\ref{fig:trimlines}c, hatched areas).

    In the upper Rhone valley, the geographical location of observed trimlines
    show remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based for less than 1\,ka (7\,\%) of the
    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
    detailed comparisons across the
    entire Alpine range and specific sensitivity studies to relevant basal
    sliding and ice rheological parameters, the presence of an upper layer of
    cold ice during the LGM, already found in some places in a much warmer
    climate today \todo{ref.}, is inevitable. Thus we want to challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer cold ice in the Alps.

    \todo{compute relative difference.}

@@

    As previously mentioned (Sect.~\ref{sec:timeseries}), independently of the
    palaeo-temperature (GRIP, EPICA, or MD01-2444) and palaeo-precipitation
    (with or without corrections) applied, all simulations presented here
    result in a high time variability in the total modelled ice volume
    (Fig.~\ref{fig:timeseries}). Using the optimal EPICA palaeo-temperature
    record with least variability (Sect.~\ref{sec:paltemp}) and conservatively
    including palaeo-precipitation reductions (Sect.~\ref{sec:palprec}), the
    1-km resolution simulation also results in strong total ice volume
    variability throughout the last glacial cycle (Fig.~\ref{fig:lgmvel}b).

    Two major glaciations occur during MIS~4 and 2 (Fig.~\ref{fig:lgmvel}b).
    However, several minor glaciations occur during MIS~5 and 3, as well as a
    minor late-glacial readvance at the onset of MIS~1
    (Fig.~\ref{fig:lgmvel}b). These episodes are the result of synchronous
    advances of several Alpine glaciers well into the major valleys and sometimes
    even onto the foreland (Fig.~\ref{fig:profiles}), Supplementary animation).

@@

    Despite the low temperature variability of the palaeo-climate forcing, and
    reduced precipitation dampening glacier response, our simulation depict
    the Alpine ice complex as highly dynamic, with many more than two or three
    \citep[cf.][]{Preusser.2004,Ivy-Ochs.etal.2008} glaciations, and regional
    glacier dynamic controlled by variations in subglacial topography and
    catchment geometry.
